---
title: "Data Preparation,Analysis and Learning in R"
author: "Juweria Ali"
date: "25/02/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Clearing the workspace and setting the working directory.

```{r warning = F}
rm(list=ls())
```
# Set working directory
```{r}
#library(here)
#setwd(here::here())
```
# Loading libraries
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(caret)
library(Hmisc)
library(ggplot2)
library(rattle)
library(xgboost)
```
# Loading & reading the datasets - Analysis and Escapes 
```{r}
Escapes <- read.csv("escapes.csv", header=T, stringsAsFactors=F)
Analysis <- read.csv("analysis.csv", header=T, stringsAsFactors=T)
```
#Q1) Preparing the data sets
##Removing unnecessary columns
```{r}
Escapes$Aquaculture.Type <- NULL # Removing Aquaculture Type, as it has no variation (i.e. factors with one level)

Escapes$Escape.Grid.Reference <- NULL # Removing Escape Grid Reference, as reference points are not useful in analysis

Escapes$Escape.ID <- NULL #Removing Escape ID,  the value is unique to all the instances and does not add value to the overall analysis and can be deleted

Escapes$Escape.Water.Type <- NULL #This column is an abbreviation of the column Water.type, hence not required
Escapes$Marine.Scotland.Site.ID <-NULL
Escapes$Initial.Escape.Reason <-NULL
Escapes$National.Grid.Reference <-NULL
Escapes$Species <- NULL
Escapes$Easting <-NULL
Escapes$Northing <-NULL
Escapes$Site.Address.1 <-NULL
Escapes$Site.Address.2 <-NULL
Escapes$Site.Address.3 <-NULL
Escapes$Escape.Start.Time <-NULL
Escapes$MS.Management.Area <-NULL
Escapes$Date.Registered <-NULL
Escapes$Producing.in.Last.3.Years <-NULL
```
##Cleaning up column Age
##Replacing 'unknown' with NA
```{r}
Escapes$Age[Escapes$Age == "unknown"] <- NA 
```
##Replacing the instances with a string value to NA's
```{r}
Escapes$Age[Escapes$Age == "unknown"]<-NA
Escapes$Age[Escapes$Age == "post smolt"]<-NA
Escapes$Age[Escapes$Age == "parr/presmt"]<-NA
Escapes$Age[Escapes$Age == "not report"]<-NA
Escapes$Age[Escapes$Age == "parr"]<-NA
Escapes$Age[Escapes$Age == "parr/presm"]<-NA

```
##Extracting the numeric values from the column age 
```{r}
Escapes$Age01<- str_extract_all(Escapes$Age, "[0-9]+") 
head(Escapes$Age01)
```
##Stripping value of instances from the new column into three columns namely New_Age
```{r}
Escapes1<- Escapes%>% separate(Age01, into = c("New_Age"), sep = "_", remove = TRUE) 
```
##Removing columns Age,Age02 and Age03 as they are no longer useful
```{r}
Escapes2<-subset(Escapes1,select=-c(Age)) 
```
##There are instances in the column that are a list.The function below identifies and replaces/imputes them with NA's
```{r}
Escapes2$New_Age<-as.numeric(unlist(Escapes2$New_Age))
```
##Histogram to check for any outliers
```{r}
hist(Escapes2$New_Age)
```
##Outliers noticed, deal with them
```{r}
Escapes2$New_Age[Escapes2$New_Age == 1999] <-NA
Escapes2$New_Age[Escapes2$New_Age == 2000] <-NA
```
##Imputing NA's in the column New_Age with median
```{r}
Escapes2$New_Age <- impute((Escapes2$New_Age), median)
```
##Cleaning the column Average Weight column
##Replacing the instances with string values i.e. unknown and post smolt to NA's
```{r}
Escapes2$Average.Weight[Escapes2$Average.Weight == "unknown"]<-NA
Escapes2$Average.Weight[Escapes2$Average.Weight == "post smolt"]<-NA
```
##We only need the numeric values from the Average.Weight column. In order to do this we separate the column into kg and g column,then, filter the columns as shown below.
```{r}
Escapes3 <- Escapes2%>% filter(grepl('kg|kilos', Average.Weight))  %>% filter(!grepl('-', Average.Weight))

Escapes4 <- Escapes2%>% filter(!grepl('k', Average.Weight)) %>% filter(grepl('g', Average.Weight))  %>% filter(!grepl('-', Average.Weight))
```
##As there are some instances that do not have a kg or g,now we merge the two columns created above and get the difference of what was left after separating the kg and g.
```{r}
Escapes_join<-merge(Escapes3,Escapes4,all=TRUE)
difference <- setdiff(Escapes2,Escapes_join)
```
##Substituting the grams and kilograms with blank spaces and converting the grams column to kilogram by dividing with 1000.
```{r}
Escapes3<-Escapes3 %>%  mutate(Average.Weight=as.double(gsub('kg','',as.character(Escapes3$Average.Weight))))

Escapes4<- Escapes4%>%  mutate(Average.Weight=as.double(gsub('g','',as.character(Escapes4$Average.Weight))))

Escapes4$Average.Weight<- Escapes4$Average.Weight / 1000
difference$Average.Weight <- NA
```
##Merging all the three above data frames back to one data frame Escapes6
```{r}
Escapes5 <- union(Escapes3,Escapes4) 
Escapes6 <- union(Escapes5,difference)
```
## Imputing NA's in the column Average Weight with median
```{r}
Escapes6$Average.Weight <- impute((Escapes6$Average.Weight), median)
Escapes6$Average.Weight<-as.integer(Escapes6$Average.Weight)
boxplot(Escapes6$Average.Weight, ylab = "Weight")
```
#Preparing the datasets fro merge
##Extracting the numeric values from the colummns Initial Number Escaped,Final Number Escaped,Final Number Recovered
```{r}
Escapes6$Initial.Number.Escaped <- as.integer(str_extract(Escapes6$Initial.Number.Escaped, "[0-9]+"))
Escapes6$Final.Number.Escaped <- as.integer(str_extract(Escapes6$Final.Number.Escaped, "[0-9]+"))
Escapes6$Final.Number.Recovered <- as.integer(str_extract(Escapes6$Final.Number.Recovered, "[0-9]+"))
```
##Creating a new dataframe with only the columns Initial.Number Escaped,Final.Number.Escaped, Final.Number.Recovered,New_Age,Average.Weight to apply bag impute
```{r}
Escapes7<-Escapes6 %>% select(Initial.Number.Escaped, Final.Number.Escaped, Final.Number.Recovered,New_Age,Average.Weight)
```
##Bag Impute applied to replace NA's 
```{r}
bagImp <- preProcess(Escapes7, method = c("bagImpute"))
bagimputedEscapes <- predict(bagImp, Escapes7)
```
##Creating a new dataframe by deleting the columns Initial.Number.Escaped,Final.Number.Escaped,Final.Number.Recovered,New_Age,Average.Weight in order to merge it with the dataframe with bag imputed values
```{r}
EscapesnoCol<-subset(Escapes6,select=-c(Initial.Number.Escaped, Final.Number.Escaped, Final.Number.Recovered,New_Age,Average.Weight))
```
##Combining the dataframe with EscapesnoCol and bagimputedEscapes, assigning that to a new dataframe EscapesCol, then assigning that new dataframe to EscapesFinal
```{r}
EscapesFinal<-bind_cols(EscapesnoCol,bagimputedEscapes)
```
#Removing unnecessary columns
```{r}
EscapesFinal$Initial.Number.Escaped <-NULL
EscapesFinal$Final.Number.Recovered <- NULL
EscapesFinal$Initial.Date.of.Escape <- NULL
EscapesFinal$Final.Date.of.Escape <- NULL
EscapesFinal$Escape.End.Time <- NULL
```
##Formatting the column Escape start date
###the dates are  in  character format, here we are parsing them in a date format as 'dmy'
```{r}
EscapesFinal$Escape.Start.Date <-
  dmy(EscapesFinal$Escape.Start.Date) 
```
##Extracting month and year as new columns
```{r}
EscapesFinal$StartingMonth <- as.character(as.integer(format(EscapesFinal$Escape.Start.Date, "%m")))
EscapesFinal$StartingYear<- as.character(as.integer(format(EscapesFinal$Escape.Start.Date, "%Y")))
```
##Changing site name in both datasets to lower case in order make them uniform
```{r}
EscapesFinal$Site.Name <- tolower(EscapesFinal$Site.Name)
Analysis$Site.Name <- tolower(Analysis$Site.Name)
```
##Converting the month and year columns of the Analysis dataset to character type so to match the EscapesFinal dataset column type
```{r}
Analysis$month <- as.character(Analysis$month)
Analysis$year<- as.character(Analysis$year)
```
##Creating a new column Grp in the Analysis dataframe by combining month,year and Site.Name,to then compare and see if there are any duplicates and then everything that is not a duplicate stays in Analysis dataframe.
```{r}
Analysis$Grp <- paste(Analysis$month,Analysis$year,Analysis$Site.Name)
Analysis<-Analysis[!duplicated(Analysis$Grp), ]
```
##Deleting the column Grp as it is no longer needed
```{r}
AnalysisFinal <- subset(Analysis, select = -c(Grp))
```
##Removing other columns in the Analysis dataset that are not useful
```{r}
AnalysisFinal <- subset(AnalysisFinal, select = -c(c2,c3,c4,c5,c6,c7))
```
#Q2)Integrate the 2 datasets together into a merged dataset
## EscapesFinal and AnalysisFinal datasets
```{r}
escapesPlus <- EscapesFinal %>% left_join(AnalysisFinal, AnalysisFinal, by=c('Site.Name'='Site.Name', 'StartingYear'='year','StartingMonth'='month'))

write.csv(escapesPlus, "escapesPlus.csv")
```
##Summary of the new merged dataset
```{r}
summary(escapesPlus)
```
#Q3)Exploratory data analysis of the dataset
##Univariate Analysis
##Checking class distribution for Escaped.Sepcies,Final.Escape.Reason in order to identify any interesting insights.
##Plotting a bar chart on the Escaped.Species to check the class distribution.
```{r}
plot1 <- ggplot(escapesPlus, aes(x=reorder(Escaped.Species, Escaped.Species, function(x)-length(x)))) +
geom_bar(fill='Green') +  labs(x='Species')
plot1
```
##Average Weight is a continuous variable
###Spread
```{r}
summary(escapesPlus$Average.Weight)
hist(escapesPlus$Average.Weight, main = "Spread for Weight",col = "#1b98e0")
                         
```
##New_Age is a continuous variable
###Spread
```{r}
summary(escapesPlus$New_Age)
hist(escapesPlus$New_Age, main = "Spread for Age",col = "#1b98e0")
                         
```
#Bivariate analysis to see impact of New_Age on Escaped Species
##stacked bar chart
```{r}
ggplot(escapesPlus,
       aes(x = New_Age,
           fill = Escaped.Species)) +
  geom_bar(position = "stack")+
labs(title = "Stack bar chart for Age and Escaped Species",x="New_Age",y="Escaped.Species")
```
#Bivariate analysis to see impact of Average weight on Escaped Species
##stacked bar chart
```{r}
ggplot(escapesPlus,
       aes(x = Average.Weight,
           fill = Escaped.Species)) +
  geom_bar(position = "stack")+
labs(title = "Stack bar chart for Average Weight and Escaped Species",x="Average Weight",y="Escaped.Species")
```
## Checking for any correlation between the age of the fish and their weight
```{r}
escapesPlus$New_Age <- as.integer(as.character(escapesPlus$New_Age))
byEscapeSpecies <- group_by(escapesPlus, Escaped.Species)
groupedDetails <- summarise(byEscapeSpecies,
                    count = n(),
                    averageAge = mean(New_Age, na.rm=T),
                    medianAge = median(New_Age, na.rm=T), 
                    oldest = max(New_Age, na.rm=T),
                )
groupedDetails
```

```{r}
escapesPlus$Average.Weight <- as.integer(as.character(escapesPlus$Average.Weight))
byEscapeSpecies <- group_by(escapesPlus, Escaped.Species)
groupedDetails <- summarise(byEscapeSpecies,
                    count = n(),
                    averageWeight = mean(Average.Weight, na.rm=T),
                    medianWeight = median(Average.Weight, na.rm=T), 
                    heavy = max(Average.Weight, na.rm=T),
                   
                    )
groupedDetails
```
#Q4)Preparing the dataset for learning
##We want to predict the feature Average.Weight, using the rest of the data in the NewEscapesPlus dataset
## Selecting columns that are useful for learning
```{r}
escapesPlus <- subset(escapesPlus, select = c(Escape.Start.Date,Average.Weight,New_Age,Escaped.Species,Stage,Final.Escape.Reason))
```
##Selecting instnaces with only atlantic salmon and rainbow trout as species
```{r}
NewEscapesPlus <- escapesPlus[ which( escapesPlus$Escaped.Species=="atlantic salmon"| escapesPlus$Escaped.Species == "rainbow trout") , ]
```
##Removing missing values if any
```{r}
NewEscapesPlus = NewEscapesPlus[complete.cases(NewEscapesPlus), ]
```
##Applying learning tasks
##Categorical target with rpart, rf and xgboost
###Setting train controls (repeated cv)
```{r}
trControl2 <- trainControl(method = "repeatedcv", number=10, repeats=1)
```
##Decision tree model
```{r}
set.seed(1234)
rpart.model <- train(Escaped.Species~.,
    data = NewEscapesPlus,
    method = "rpart",
    metric = "Accuracy",
    trControl = trControl2)

print(rpart.model)
confusionMatrix.train(rpart.model)
fancyRpartPlot(rpart.model$finalModel)
```
##Random forest
```{r}
set.seed(1234)
rf.model <- train(Escaped.Species~.,
    data = NewEscapesPlus,
    method = "rf",
    metric = "Accuracy",
    trControl = trControl2)
print(rf.model)
confusionMatrix.train(rf.model)

```
##Boosted trees
```{r}
set.seed(1234)
# Run the model

xgboost.model <- train(Escaped.Species ~ ., 
                       data=NewEscapesPlus,   
                       metric = "Accuracy",
                       method="xgbTree",
                       trControl = trControl2)
#print(xgboost.model)
confusionMatrix.train(xgboost.model)
```
#Comparing results of the three models
```{r}
results <- resamples(list(CART=rpart.model, randomForest = rf.model, xgboost = xgboost.model))
results
summary(results)

```
#To plot the results
```{r}
scales <- list(x=list(relation="free"), y=list(relation= "free"))
dotplot(results, scales=scales, conf.level = 0.95)

```
#Creating train and test sets

```{r}
set.seed(1234)
partIndex <- createDataPartition(NewEscapesPlus$Average.Weight, p=0.75, list=F)
trainData <- NewEscapesPlus[partIndex,]
testData <- NewEscapesPlus[-partIndex,]

```
#Training rpart
```{r}
set.seed(1234)
rpart.model.tr <- train(Average.Weight ~ ., data=trainData, 
                    method="rpart",
    trControl = trControl2, na.action=na.omit)
print(rpart.model.tr)
```
#Testing rpart
```{r}
rpartRes <- predict(rpart.model.tr, newdata = testData)
table(rpartRes, testData$Average.Weight)
```
#Training random forest
```{r}
set.seed(1234)
rf.model.tr <- train(Average.Weight ~ ., data=trainData, 
                    method="rf",
    trControl = trControl2, na.action=na.omit)
print(rf.model.tr)
```
#Testing random forest
```{r}
rfRes <- predict(rf.model.tr, newdata = testData)
table(rfRes, testData$Average.Weight)
```
#Training boosted tree
```{r}
set.seed(1234)
xgb.model.tr <- train(Average.Weight ~ ., data=trainData, 
                    method="xgbTree",
    trControl = trControl2, na.action=na.omit)
print(xgb.model.tr)
```
#Testing boosted tree
```{r}
xgbRes <- predict(xgb.model.tr, newdata = testData)
table(xgbRes, testData$Average.Weight)
```
#Comparing results
```{r}
results <- resamples(list(CART=rf.model.tr, randomForest = rf.model.tr, xgboost = xgb.model.tr))
results
summary(results)
```





















